---
title: "Introduction to dyntaxa"
author: "Markus Skyttner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dyntaxa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

Start by loading the package and download the Dyntaxa database for local use.

```{r eval=FALSE}
library(dyntaxa)

# the first time the package is loaded, it attempts to
# download and generate Dyntaxa data assets locally for use with the package
# and only after this, the various functions can then be used to query the Dyntaxa database
# this happens in the dyntaxa_init() function, which can be forced to download 
# and overwrite the locally persisted data in order to refresh it, like so:
dyntaxa_init_fts()

```

The data is available locally after the initial download through the `dyntaxa_init_fts()` function, which will requires an internet connection in other to retrieve a remote Darwin Core Archive file with open data from the Dyntaxa database. After this initial download of the data, the package can be used off-line without Internet connectivity.

## Basic usage

Basic usage examples follow, showing how to look up taxonomic data from the Dyntaxa database related to specific (single) taxonomic names or identifiers. The locally stored data has a full text search index generated, which can be used for taxonomic full text search queries.


```{r basic-example}
library(dyntaxa)
library(dplyr)

# taxonomic name from taxonomic identifier
# NB: quoting of the string
dyntaxa_search('"urn:lsid:dyntaxa.se:Taxon:220023"') %>% pull(scientificName)

# taxonomic identifier from taxonomic name
# NB: usage of + to match both search terms
key <- dyntaxa_search("Alces+alces") %>% pull(taxonId)
key

# taxonomic hierarchy/classification from identifier or name
dyntaxa_search(paste0('"', key, '"')) %>%
  select(kingdom, phylum, class, order, family, genus, species) %>%
  glimpse()

# taxa downstream from identifier or name
dyntaxa_search("Cervidae") %>% filter(taxonomicStatus == "accepted") %>%
  collect()

# taxonomic immediate children of an identifier or name
dyntaxa_search("Carnivora") %>% filter(taxonRank == "suborder") %>%
  filter(taxonomicStatus == "accepted") %>%
  collect()

# taxa downstream of a specific taxon identifier
# can be filtered at species level (or at other ranks)
dyntaxa_search("Carnivora") %>% filter(taxonRank == "species") %>%
  filter(taxonomicStatus == "accepted") %>%
  collect()

# what are synonyms for Sagedia zonata?
dyntaxa_search("Sagedia+zonata synonym") %>% collect() %>% pull(scientificName)

# does Citronfjäril have synonyms?
dyntaxa_search("citronfjäril") %>%
  select(taxonomicStatus, scientificName)

```

## Automating lookups for lists with taxonomic names and/or identifiers

In general, when working with lists (or vectors) of several species names (or identifiers), using packages such as `purrr` and `dplyr` will enable convenient automated lookups.

The following usage examples shows how to automate a number of tasks for variable-length lists of taxonomic names or identifiers and return results as neat data tables (tibbles):

```{r lists-example, eval=TRUE}

library(purrr)
library(dplyr)

# given a list of several species names, lookup the identifiers
lookups <- c("Abies procera", "Pinus contorta")
df <- dyntaxa_search() %>% filter(scientificName %in% lookups)
df

# the opposite - given a list of several identifiers, lookup the names
keys <- df$taxonId
dyntaxa_search() %>% filter(taxonId %in% keys)

# search fulltext index for several terms
dyntaxa_search("blåklocka OR vitsippa") %>% 
  select(taxonId, scientificName, vern)

```
